Using Google Community Mobility Reports to investigate the incidence of COVID-19 in the United States

### 2020-05-19 Covid-19-Google-CMR-US/Response-to-Reviewers-v3.Rmd

---
title: "Response to Reviewers"
author: "Antonio Paez"
date: "5/11/2020"
output: pdf_document
---

```{r load-packages, include=FALSE}
#Load the packages needed to read and work with the data
library(adespatial)
library(countrycode)
library(gridExtra)
library(interactions)
library(jtools)
library(kableExtra)
library(lubridate)
library(metR)
library(plotly)
library(RCurl)
library(sf)
library(spdep)
library(tidyverse)
library(tigris)
library(usmap)
library(wppExplorer)
```

```{r load-saved-gcmr-data, include=FALSE}
#Load data:
load("gmobility.RData")
```

```{r extract-gcmr-for-us, include=FALSE}
gmobility <- gmobility %>%
  filter(country_region == "United States") %>%
  droplevels()
```

```{r prepare-mobility-report-data, include=FALSE}
#`date` is stored as a factor which is not ideal work with times; Change format:
gmobility <- gmobility %>%
  mutate(date = lubridate::ymd(date)) %>%
  #rename the annoyingly long names of the columns to something more practical
  transmute(date,
            state = as.character(sub_region_1),
         county = as.character(sub_region_2),
         r_and_r = retail_and_recreation_percent_change_from_baseline, 
         g_and_p = grocery_and_pharmacy_percent_change_from_baseline, 
         parks = parks_percent_change_from_baseline,
         transit = transit_stations_percent_change_from_baseline, 
         work = workplaces_percent_change_from_baseline,
         residential = residential_percent_change_from_baseline) %>%
  #convert the mobility indicators to differences from the baseline, so "1" is the baseline value
  mutate(r_and_r = 1 + r_and_r/100,
         g_and_p = 1 + g_and_p/100,
         parks = 1 + parks/100,
         transit = 1 + transit/100,
         work = 1 + work/100,
         residential = 1 + residential/100)
```

```{r filter-state-gcmr-data, include=FALSE}
#To extract the state-level mobility indicators select those records that do not include the label for the county (i.e., sub_region_2 is empty), and that are not national level statistics (i.e., sub_region_1 is not empty)
cmr <- gmobility %>%
  filter(state !="", 
         county == "")
```

```{r load-saved-covid-19-data, include=FALSE}
#Load data:
load("covid19_nyt.RData")
```

```{r coterminous-states, include=FALSE}
#filter states to keep only the coterminous states
c19 <- c19 %>%
  filter(state != "Hawaii",
         state != "Alaska",
         state != "United States Virgin Islands",
         state != "Commonwealth of the Northern Mariana Islands",
         state != "Guam",
         state != "American Samoa",
         state != "Puerto Rico")
```

```{r add-day-variable, include=FALSE}
#add day variable:
c19 <- c19 %>%
  mutate(day = yday(date) - yday("2020-04-05"))
```

```{r calculate-new-daily-cases, include=FALSE}
#To calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total. Also calculate the daily percent change in new cases (pct_change), and the percent increase in total cases (pct_increase):
c19 <- c19 %>%
  group_by(state) %>%
  mutate(new_cases = cases - lag(cases, 1),
         pct_change = (new_cases - lag(new_cases, 1))/lag(new_cases, 1),
         pct_increase = (cases - lag(cases, 1))/lag(cases, 1)) %>%
  ungroup() %>%
  #there are two record with negative numbers of new cases, likely corrections (Georgia in 2020-04-12, Idaho in 2020-05-03). Remove record:
  filter(new_cases >= 0)
```

```{r state-population-data, include=FALSE}
#Retrieve population data:
population <- statepop %>%
  transmute(pop_2015, state = full)
```

```{r join-population-to-covid-19-data, include=FALSE}
#Join population to covid-19 data:
c19 <- c19 %>%
  mutate(state = as.character(state)) %>%
  left_join(population, by = c("state" = "state"))
```

```{r calculate-incidence, include=FALSE}
#Calculate the incidence of reported cases by 100,000 population:
c19 <- c19 %>%
  mutate(incidence = cases/(pop_2015/100000),
         log_incidence = log(incidence))
```

```{r calculate-log-new-cases, include=FALSE}
#calculate log of new cases
c19 <- c19 %>% 
  mutate(log_new_cases = log(new_cases + 0.001))
```

```{r join-gcmr-and-covid-19-data, include=FALSE}
#Join the tables:
c19 <- c19 %>% 
  mutate(state = as.character(state)) %>%
  left_join(cmr, by = c("state" = "state", "date" = "date"))
```

```{r calculate-lagged-mobility-indicators, include=FALSE}
#The potential salutary effect of the lockdown (measured in terms of reduced mobility) are not immediate, given the incubation period of the disease. The incubation period has been estimated to be between 2 and 11 days (95% interval) by [Lauer et al. (2020)](https://annals.org/aim/fullarticle/2762808/incubation-period-coronavirus-disease-2019-covid-19-from-publicly-reported). If we wanted to explore the way reduced mobility correlates with number of new cases, we would need to lag the mobility variables by some period.

# Lagged retail and recreation: 11 day moving average
r_and_r_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$r_and_r, 12) + 
                      lag(.x$r_and_r, 11) + 
                      lag(.x$r_and_r, 10) + 
                      lag(.x$r_and_r, 9) + 
                      lag(.x$r_and_r, 8) + 
                      lag(.x$r_and_r, 7) + 
                      lag(.x$r_and_r, 6) + 
                      lag(.x$r_and_r, 5) + 
                      lag(.x$r_and_r, 4) + 
                      lag(.x$r_and_r, 3) + 
                      lag(.x$r_and_r, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_r_and_r_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_r_and_r_lag11)

# Lagged groceries and pharmacies: 11 day moving average
g_and_p_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$g_and_p, 12) + 
                      lag(.x$g_and_p, 11) + 
                      lag(.x$g_and_p, 10) + 
                      lag(.x$g_and_p, 9) + 
                      lag(.x$g_and_p, 8) + 
                      lag(.x$g_and_p, 7) + 
                      lag(.x$g_and_p, 6) + 
                      lag(.x$g_and_p, 5) + 
                      lag(.x$g_and_p, 4) + 
                      lag(.x$g_and_p, 3) + 
                      lag(.x$g_and_p, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_g_and_p_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_g_and_p_lag11)

# Lagged parks: 11 day moving average
parks_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$parks, 12) + 
                      lag(.x$parks, 11) + 
                      lag(.x$parks, 10) + 
                      lag(.x$parks, 9) + 
                      lag(.x$parks, 8) + 
                      lag(.x$parks, 7) + 
                      lag(.x$parks, 6) + 
                      lag(.x$parks, 5) + 
                      lag(.x$parks, 4) + 
                      lag(.x$parks, 3) + 
                      lag(.x$parks, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_parks_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_parks_lag11)

# Lagged groceries and pharmacies: 11 day moving average
transit_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$transit, 12) + 
                      lag(.x$transit, 11) + 
                      lag(.x$transit, 10) + 
                      lag(.x$transit, 9) + 
                      lag(.x$transit, 8) + 
                      lag(.x$transit, 7) + 
                      lag(.x$transit, 6) + 
                      lag(.x$transit, 5) + 
                      lag(.x$transit, 4) + 
                      lag(.x$transit, 3) + 
                      lag(.x$transit, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_transit_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_transit_lag11)

# Lagged groceries and pharmacies: 11 day moving average
work_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$work, 12) + 
                      lag(.x$work, 11) + 
                      lag(.x$work, 10) + 
                      lag(.x$work, 9) + 
                      lag(.x$work, 8) + 
                      lag(.x$work, 7) + 
                      lag(.x$work, 6) + 
                      lag(.x$work, 5) + 
                      lag(.x$work, 4) + 
                      lag(.x$work, 3) + 
                      lag(.x$work, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_work_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_work_lag11)

# Lagged residential: 11 day moving average
residential_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$residential, 12) + 
                      lag(.x$residential, 11) + 
                      lag(.x$residential, 10) + 
                      lag(.x$residential, 9) + 
                      lag(.x$residential, 8) + 
                      lag(.x$residential, 7) + 
                      lag(.x$residential, 6) + 
                      lag(.x$residential, 5) + 
                      lag(.x$residential, 4) + 
                      lag(.x$residential, 3) + 
                      lag(.x$residential, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_residential_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_residential_lag11)
```

```{r join-lagged-indicators, include=FALSE}
#join lagged mobility indicators to table:
c19 <- c19 %>%
  mutate(state = as.character(state)) %>%
  left_join(r_and_r_lag11, by = c("state" = "state", "date" = "date")) %>%
  left_join(g_and_p_lag11, by = c("state" = "state", "date" = "date"))  %>%
  left_join(parks_lag11, by = c("state" = "state", "date")) %>%
  left_join(transit_lag11, by = c("state" = "state", "date")) %>%
  left_join(work_lag11, by = c("state" = "state", "date")) %>%
  left_join(residential_lag11, by = c("state" = "state", "date"))
```

```{r remove-nas, include=FALSE}
#Remove all the "NAs" to obtain a clean table with COVID-19 and mobility information:
c19 <- c19 %>%
  drop_na()
```

```{r prepare-variables-for-analysis, include=FALSE}
#replace mobility indicators by their lagged moving averages
c19 <- c19 %>% 
  mutate(retail = Mean_r_and_r_lag11,
         groceries = Mean_g_and_p_lag11,
         parks = Mean_parks_lag11,
         parks2 = Mean_parks_lag11^2,
         transit = Mean_transit_lag11,
         work = Mean_work_lag11,
         work2 = work^2,
         residential = Mean_residential_lag11,
         residential2 = Mean_residential_lag11^2,
         parksXresidential = parks * residential,
         parksXwork = parks * work,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks * day2,
         workXday = work * day,
         workXday2 = work * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NY = ifelse(state == "New York", 1, 0),
         NYxday = NY * day)
```

# Reviewer 1

Thanks for the revisions. I think they make the paper stronger.

>\textcolor{blue}{Thank you for your positive assessment of the paper and comments.}

I have a question on the rationale for using daily new cases (log transformed) as a dependent variable. Given the large differences in population size, urban areas and densities across states, wouldn’t using percentage change in new cases day to day be a more appropriate dependent variable (or some moving average of it)? Such a setup removes the population size effect...I would suggest the authors at least test the percent variable and see if it allows for a simpler specification. Or alternately, explain why log(new cases) is a valid measure to answer their question given differences across states that are not accounted for.

>\textcolor{blue}{Thank you for this comment. I understand percentage change in new cases for state $i$ and day $t$ to mean the following:}

$$
pct\_change_{i,t} = \frac{new\_cases_{i,t} - new\_cases_{i,t-1}}{new\_cases_{i,t-1}}
$$

>\textcolor{blue}{Alas, as seen in Table \ref{tab:correlation-analysis-pct-change} below, percent change correlates poorly with the independent variables, and the low correlations translate into a poor fit with non-significant coefficients (see Table \ref{tab:model-results-pct-change}). This is because the daily percent change in new cases is very noisy and not a good indication of a trend: even small changes (say from +1 to -1 case change from day to day) change the sign of the variable.}

```{r check-correlations-pct-change, echo=FALSE}
#Check the correlation between the mobility indicators and day-to-day percent change in new cases:
c19 %>% 
  filter(is.finite(pct_change)) %>%
  select(pct_change, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis-pct-change}Simple correlation between day to day percent change in new Cases and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

```{r model-estimation-pct-change, include=FALSE}
mod_pct_change <- lm(pct_change ~  0 +
                         day +
                         day2 +
                         parks2 + 
                         parks + 
                         parksXwork +
                         work +
                         work2 +
                         parksXday + 
                         parksXday2 +
                         workXday +
                         workXday2 +
                         NY + 
                         NYxday, 
                       data = c19 %>% filter(is.finite(pct_change))
)
summ(mod_pct_change)
```

```{r model-results-pct-change, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x work",
                        "work",
                        "work^2",
                        "parks x date",
                        "parks x date^2",
                        "work x day^2",
                        "work x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod_pct_change)),
           pval = as.numeric(summary(mod_pct_change)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results-pct-change}Dependent variable is day-to-day percent change in new cases.",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod_pct_change)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod_pct_change)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod_pct_change)$sigma, 3))),
           escape = FALSE)
```

>\textcolor{blue}{Instead of day-to-day percent change in new cases, we could define a variable that measures the percentage increase in total cases:}

$$
pct\_increase_{i,t} = \frac{cases_{i,t} - cases_{i,t-1}}{cases_{i,t-1}}
$$

>\textcolor{blue}{Again, a variable like this would normalize for the caseload in each state. The correlations for percent increase are shown in Table \ref{tab:correlation-analysis-pct-increase}, were we see that they are considerably higher than for day-to-day percent change in cases, but still lower than for log of new cases. If we use daily percent increase in total cases as the dependent variable, the resulting model is somewhat more parsimonious but the goodness of fit is also quite a bit lower than the model with log of new cases (see Table \ref{tab:model-results-pct-increase}).}

```{r check-correlations-pct-increase, echo=FALSE}
#Check the correlation between the mobility indicators and daily percent increase in total cases:
c19 %>% 
  select(pct_increase, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis-pct-increase}Simple correlation between daily percent increase in total cases and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

```{r model-estimation-pct-increase, include=FALSE}
mod_pct_increase <- lm(pct_increase ~  0 +
                         #day +
                         #day2 +
                         #parks2 + 
                         parks + 
                         parksXwork +
                         work +
                         work2 +
                         parksXday + 
                         parksXday2 +
                         #workXday +
                         workXday2 +
                         NY + 
                         NYxday, 
                       data = c19)
summ(mod_pct_increase)
```

### 2020-05-19 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.Rmd
v3 Revisions after second round of anonymous reviews.

---
title: Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
   In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate potential correlations between mobility and transmission of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

journal: "Transportation Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
always_allow_html: true
---

@@

```{r calculate-new-daily-cases, include=FALSE}
#o calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total:
c19 <- c19 %>%
  group_by(state) %>%
  mutate(new_cases = cases - lag(cases, 1)) %>%
  ungroup() %>%
  #there is one record with a negative number of new cases, likely a correction (Georgia in 2020-04-12). Remove record:
  filter(new_cases >= 0)
```

@@

The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders, which in the United States have been implemented on a state-by-state basis, with considerable variations in compliance. Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic. As a result, there are new sources of data such as the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}. These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

This paper investigates the potential of Google Community Mobility Reports to asses the impact of mobility on COVID-19. The following questions are posed:

- Do changes in mobility according to GCMR correlate with the transmission of COVID-19?
- And if so, what do we learn about mobility and the spread of the disease?

@@

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places (see Table \ref{tab:descriptive-statistics}). Mobility indicators are calculated based on the frequency and length of visits. The reports give percentage change from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("New Cases",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("New Daily Cases of COVID-19",
                        "Date",
                        "Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters",
                        "Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies",
                        "Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens",
                        "Mobility trends for places like public transport hubs such as subway, bus, and train stations",
                        "Mobility trends for places of work",
                        "Mobility trends for places of residence"),
           min = c(as.character(min(c19$new_cases)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(median(c19$new_cases)),
                   as.character(median(c19$date)),
                   as.character(round(median(c19$retail), 2)),
                   as.character(round(median(c19$groceries), 2)),
                   as.character(round(median(c19$parks), 2)),
                   as.character(round(median(c19$transit), 2)),
                   as.character(round(median(c19$work), 2)),
                   as.character(round(median(c19$residential), 2))),
           max = c(as.character(max(c19$new_cases)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$new_cases), 2)),
                   as.character(""),
                   as.character(round(sd(c19$retail), 2)),
                   as.character(round(sd(c19$groceries), 2)),
                   as.character(round(sd(c19$parks), 2)),
                   as.character(round(sd(c19$transit), 2)),
                   as.character(round(sd(c19$work), 2)),
                   as.character(round(sd(c19$residential), 2)))) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
  column_spec(2, width = "12em") %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease is between 2 and 12 days [95% interval; see @Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect (if any) on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated. This variable (log-transformed after adding a small constant) is paired with the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of daily new cases when making predictions. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Analysis is based on correlation analysis, multivariate regression, and data visualization.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other. Two variables are selected for multivariate analysis: parks- and work-related mobility. Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables. Furthermore, parks- and work-related mobility represent two dimensions of out-of-home activities: mandatory and discretionary travel.

```{r check-correlations, echo=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model is estimated with the log of new daily cases as the dependent variable. The covariates enter the regression in the form of a second order polynomial expansion. In addition, the date (centered on April 5) is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there. The results appear in Table \ref{tab:model-results}. The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.

There is an overall temporal trend that indicates a growing number of cases over time, at an accelerating rate (see positive sign of date^2). Mobility related to parks and to work both tend to increase the number of new cases; the negative sign for their interaction is indicative of the trade-offs between these two forms of mobility and their impact on new cases. The influence of parks-related mobility was relatively weaker early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term). The opposite happens with work-related mobility, which started with a greater effect (positive sign of work x date term), but whose impact has declined over time (negative sign of work x date^2 term) New York has on average more new daily cases than the rest of the states, but this has declined over time (see negative sign of NY x date).

```{r model-estimation, include=FALSE}
#This is the regression model (forcing the regression through the intercept):
mod1 <- lm(log_new_cases ~  0 +
             day +
             day2 +
             parks2 + 
             parks + 
             parksXwork +
             work +
             work2 +
             parksXday + 
             parksXday2 +
             workXday +
             workXday2 +
             NY + 
             NYxday, 
           data = c19 #%>% filter(date >= "2020-04-01")
           )
summ(mod1)
```

>>>>

---
title: Using Google Community Mobility Reports to investigate the incidence of COVID-19 in the United States
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate potential correlations between mobility and incidence of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and incidence of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

journal: "Transport Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
always_allow_html: true
---

@@

```{r calculate-new-daily-cases, include=FALSE}
#To calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total. Also calculate the daily percent change in new cases (pct_change), and the percent increase in total cases (pct_increase):
c19 <- c19 %>%
  group_by(state) %>%
  mutate(new_cases = cases - lag(cases, 1),
         pct_change = (new_cases - lag(new_cases, 1))/lag(new_cases, 1),
         pct_increase = (cases - lag(cases, 1))/lag(cases, 1)) %>%
  ungroup() %>%
  #there are two record with negative numbers of new cases, likely corrections (Georgia in 2020-04-12, Idaho in 2020-05-03). Remove record:
  filter(new_cases >= 0)
```

@@

The main policy tool to control the spread of the COVID-19 pandemic has been restrictions to non-essential travel in the form of stay-at-home orders. In the United States, such orders have been implemented on a state-by-state basis with considerable variations in compliance. Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic. As a result, there are new sources of data such as the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}. These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

This paper investigates the potential of Google Community Mobility Reports to asses the impact of mobility on the incidence of COVID-19. The following questions are posed:

- Do changes in mobility according to GCMR correlate with the incidence of COVID-19?
- And if so, what do we learn about mobility and the spread of the disease?

@@

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places (see Table \ref{tab:descriptive-statistics}). Mobility indicators are calculated based on the frequency and length of visits to places. The reports give percentage change from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("Incidence",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("Total cases of COVID-19 divided by population (in 100,000s)",
                          "Date",
                          "Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters",
                          "Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies",
                          "Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens",
                          "Mobility trends for places like public transport hubs such as subway, bus, and train stations",
                          "Mobility trends for places of work",
                          "Mobility trends for places of residence"),
           min = c(as.character(round(min(c19$incidence), 2)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(round(median(c19$incidence), 2)),
                      as.character(median(c19$date)),
                      as.character(round(median(c19$retail), 2)),
                      as.character(round(median(c19$groceries), 2)),
                      as.character(round(median(c19$parks), 2)),
                      as.character(round(median(c19$transit), 2)),
                      as.character(round(median(c19$work), 2)),
                      as.character(round(median(c19$residential), 2))),
           max = c(as.character(round(max(c19$incidence), 2)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$incidence), 2)),
                  as.character(""),
                  as.character(round(sd(c19$retail), 2)),
                  as.character(round(sd(c19$groceries), 2)),
                  as.character(round(sd(c19$parks), 2)),
                  as.character(round(sd(c19$transit), 2)),
                  as.character(round(sd(c19$work), 2)),
                  as.character(round(sd(c19$residential), 2)))) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
  column_spec(2, width = "12em") %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

For analysis, all mobility indicators are centered so that the value of 1 is the baseline mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease is between 2 and 12 days [95% interval; see @Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the incidence is calculated after dividing by the population of the state (in 100,000s). This variable (log-transformed) is paired with the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of incidence when making predictions. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Analysis is based on correlation analysis, multivariate regression, and data visualization.

@@

Table \ref{tab:correlation-analysis-log-incidence} shows that the mobility indicators are highly correlated with each other. Two variables are selected for multivariate analysis: parks- and work-related mobility. Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables in multivariate analysis. Furthermore, parks- and work-related mobility represent two dimensions of out-of-home activities: mandatory and discretionary travel.

```{r check-correlations-log-new-cases, eval=FALSE, include=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

```{r check-correlations-log-incidence, echo=FALSE}
#Check the correlation between the mobility indicators and log of incidence:
c19 %>% 
  select(log_incidence, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis-log-incidence}Simple correlation between log(incidence) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model is estimated with the log of incidence as the dependent variable. The covariates enter the regression in the form of a second order polynomial expansion. In addition, the date (centered on April 5) is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York is used to distinguish the unusually high incidence of the disease there. The results of the model are shown in Table \ref{tab:model-results-log-incidence}. The model provides a good fit to the data and all variables reported are significant at $p<0.10$ or better.

There is an overall temporal trend that indicates a growing incidence over time, but at a decelerating rate (see negative sign of date^2). Mobility related to parks and to work are both associated with higher incidence of COVID-19, however, the effect of parks-related mobility grows non-linearly (see positive sign of quadratic term), whereas the effect of work-related mobility grows at a decreasing rate (see negative sign of quadratic term). Furthermore, the negative sign for the interaction of these two mobility indicators captures the trade-offs between these two forms of mobility and their impact on incidence. The influence of parks-related mobility was relatively weak early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term). The opposite happens with work-related mobility, the importance of which has declined over time (negative sign of work x date^2 term), but whose impact has declined over time (negative sign of work x date^2 term). As seen in the table, incidence of COVID-19 in New York is consistently higher.
